
# ggplot

When you are checking your data for distribution problems and data entry errors, quick and dirty plots are the way to go. Packages like `DataExplorer` and `skimr` have great functions to make quick exploratory plots. `ggplot` allows you more control over plot components. 

At this stage we are looking for 

1. missingness
2. weird distributions
3. values that could be errors

# set up

## load packages
```{r}
#| message: false
#| warning: false
#| 
library(tidyverse) # includes `ggplot`
library(DataExplorer)
library(skimr)

```

## read data

```{r}

penguins <- read_csv(here::here("messy_penguins.csv"))

```


# run checks

## find missingness

```{r}

plot_missing(penguins)

```


## check shape

The `skim()` function from the `skimr` package includes "sparkline" histograms in its summary stat table. `plot_histogram` from the `DataExplorer` package will plot a histogram for all numeric variables in a dataset. And if you need to look at one more closely, the base R `hist()` function is useful. 
```{r}


skim(penguins)

plot_histogram(penguins)


hist(penguins$flipper_length_mm)


```

## find errors

### categorical

To catch coding errors, plot categorical variables as a bar plot. Extra categories misspelled cells will reveal themselves. 

```{r}

plot_bar(penguins)

```


### numeric

To catch other data entry errors, plot the raw data, first as a box plot and then with jittered points overlaid. Look for impossible values. 

```{r}
#| message: false
#| warning: false

penguins %>%
  ggplot(aes(x = species, y = flipper_length_mm)) +
  geom_boxplot() 


penguins %>%
  ggplot(aes(x = species, y = flipper_length_mm)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2)


```


Making a scatterplot of variables you expect to be correlated and colouring points by another variable can make data entry errors stick out too. 

```{r}
#| message: false
#| warning: false
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, colour = species)) +
  geom_point() +
  labs(title= "A very heavy chinstrap and an Adelie with flippers that are too long")
```


> *Note `DataExplorer::create_report(df)` automagically creates a report that includes lots of data checks. They won't all be relevant to all projects, but it is a good quick and dirty way to get many of the plots you need to check your data*


# summary

| The check | The question you're asking | The plot |
|:----------|:---------------------------|:---------|
| **Missing** | What isn't here — and is it random or patterned? | `plot_missing()` / `vis_miss()` |
| **Mis-shaped** | What shape is each variable? Skew, ceiling/floor, bimodality? | `skim()`, then `plot_histogram()` |
| **Mis-coded** | Are the categories consistent, or full of typos and stray levels? | bar plot (`plot_bar`) |
| **Impossible** | Are any values out of bounds on their own? | box and jitter (`geom_boxplot()` + `geom_point()`)  |
| **Impossible together** | Are any values wrong *in combination*? | scatterplot (`geom_point`) |



: Six ways data can be wrong — and the plot that catches each {#tbl-datachecks}





